Probability

### 2016-04-27 prob.tex
prob.tex \subsection{Quantum Probability Measures} updated...

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\newcommand{\yutsung}[1]{\fbox{\begin{minipage}{0.9\textwidth}\color{purple}{Yu-Tsung says: #1}\end{minipage}}}
\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.9\textwidth}\color{green}{Amr says: #1}\end{minipage}}}
\newcommand{\ffzd}[1]{{\mathbb{F}^{d\;*}_{#1}}}
\def\C{{\mathbb{C}}}
\newcommand{\ff}[1]{\mathbb{F}_{#1}}
\newcommand{\melement}[2]{ \langle #1 | #2 | #1 \rangle}
\newcommand{\expect}[2]{ \langle #1 | #2 | #1 \rangle}
\newcommand{\Tr}{\mathop{\mathrm{Tr}}\nolimits}

@@

\begin{example}[One-qubit quantum probability space] 
  Consider a one-qubit Hilbert space with states
  $\alpha \ket{0} + \beta \ket{1}$ such that
  $|\alpha|^2 + |\beta|^2 = 1$. The set of events associated with this
  Hilbert space consists of all projection operators. Each event is
  interpreted as a possible post-measurement state of a quantum system
  in current state $\ket{\phi}$. For example, the event $\proj{0}$
  indicates that the post-measurement state will be~$\ket{0}$; the
  event $\proj{1}$ indicates that the post-measurement state will
  be~$\ket{1}$; the event $\proj{\ps}$ where
  $\ket{\ps} = \frac{1}{\sqrt{2}}(\ket{0}+\ket{1})$ indicates that the
  post-measurement state will be $\ket{\ps}$; the event
  $\mathbb{1} = \proj{0}+\proj{1}$ indicates that the post-measurement
  state will be a linear combination of $\ket{0}$ and $\ket{1}$; and
  the empty event $\mathbb{0}$ states that the post-measurement state
  will be the empty state. As in the classical case, a probability
  measure is a function that maps events to $[0,1]$: here is a partial
  specification of a possible probability measure:
\[\begin{array}{rcl}
\mu\left(\mathbb{0}\right) = 0, \quad
\mu\left(\mathbb{1}\right) =  1, \quad
\mu\left(\proj{0}\right) = 1, \quad
\mu\left(\proj{1}\right) = 0, \quad
\mu\left(\proj{\ps}\right) = 1/2, \quad \ldots
\end{array}\]
Note that, similarly to the classical case, the probability of
$\mathbb{1}$ is 1 and the probability of collections of orthogonal
events (e.g., $\proj{0}+\proj{1}$) is the sum of the individual
probabilities. In contrast, a collection of non-orthogonal events
(e.g., $\proj{0}$ and $\proj{\ps}$) is not itself an event.
In the classical example, we argued that each probability measure is
uniquely determined by two actual coins. A similar (but much more
subtle) argument is valid also in the quantum case. By postulates of
quantum mechanics and Gleason's theorem, it turns out that for large
enough quantum systems, each probability measure is uniquely
determined by an actual quantum state.
\qed\end{example}

@@

For a given set of events $\events$, there are many possible
probability measures $\mu : \events \rightarrow [0,1]$. The Born rule,
a postulate of quantum mechanics, states that each quantum state
$\ket{\phi}$ induces a probability measure $\mu_\phi$ as follows:
\[ 
\mu_\phi(E) = \ip{\phi}{E\phi}
\]
Conversely, Gleason's theorem states that given a probability measure
$\mu$, there exist a quantum state $\ket{\phi}$ that induces such a
measure using the Born rule. The theorem is only valid in Hilbert
spaces with dimension $d \geq 3$. It is instructive to study
counterexamples in $d=2$, i.e., the case of a one-qubit
system. Consider five states $\ket{\psi_0}$ to $\ket{\psi_4}$ that
form five orthogonal bases $\{ \ket{\psi_0}, \ket{\psi_1} \}$,
$\{ \ket{\psi_1}, \ket{\psi_2} \}$,
$\{ \ket{\psi_2}, \ket{\psi_3} \}$,
$\{ \ket{\psi_3}, \ket{\psi_4} \}$, and
$\{ \ket{\psi_4}, \ket{\psi_0} \}$ and consider the probability
measure defined as follows. For all $i \in \{0,1,2,3,4\}$, we have
$\mu_X(\proj{\psi_i}) = 1/2$. For each orthogonal basis, the
probability is 1 as desired and yet it is impossible to find a single
quantum state that realizes such a probability
measure (see \url{http://tph.tuwien.ac.at/~svozil/publ/2006-gleason.pdf})

@@

Although it seems that we need an infinite long table to specify the
quantum probability measure~$\mu$, our $\mu$ is actually
given by a simple formula~$\melement{0}{E}$. In general, Born discovered
each quantum state $\ket{\psi}\in\Hilb\backslash\left\{ 0\right\} $
induces a probability measure $\tilde{\mu}_{\psi}:\qevents\rightarrow[0,1]$
on the space of events defined for any event $E\in\qevents$ as follows~\cite{Born1984,Mermin2007}:
\begin{equation}
\tilde{\mu}_{\psi}(E)=\frac{\melement{\psi}{E}}{\ip{\psi}{\psi}}\label{eq:Born}
\end{equation}
The Born rule satisfies the following properties:
\begin{itemize}
\item It can be extend to mixed states. Given a mixed state represented
by a density matrix $\rho=\sum_{j=1}^{N}q_{j}\frac{\proj{\psi_{j}}}{\ip{\psi_{j}}{\psi_{j}}}$,
where $\sum_{j=1}^{N}q_{j}=1$, i.e., $\Tr\left(\rho\right)=1$, then
the Born rule can be extended to $\rho$ by 
\begin{eqnarray}
\tilde{\mu}_{\rho}\left(E\right) & = & \Tr\left(\rho E\right)=\sum_{j=1}^{N}q_{j}\tilde{\mu}_{\Psi_{j}}\left(E\right)\textrm{ .}\label{BornRule.mixed}
\end{eqnarray}
Notice that $\left(\left\{ 1,\ldots,N\right\} ,2^{\left\{ 1,\ldots,N\right\} },\pmeas\left(J\right)=\sum_{j\in J}q_{j}\right)$
is a classical probability space. Therefore, when we discretize the
Hilbert space later, we may need to discretize this probability space
as well.
\item $\tilde{\mu}_{\rho}$ is a probability measure for all mixed state~$\rho$.
\item $\ip{\psi}{\phi}=0\Leftrightarrow\tilde{\mu}_{\psi}\left(\proj{\phi}\right)=0$.
\item $\tilde{\mu}_{\psi}\left(E\right)=\tilde{\mu}_{\mathbf{U}\ket{\psi}}\left(\mathbf{U}E\mathbf{U}^{\dagger}\right)\textrm{ ,}$where
$\mathbf{U}$ is any unitary map, i.e., $\mathbf{U}^{\dagger}\mathbf{U}=\mathbb{1}$. 
\end{itemize}

Naturally, we may ask: is every probability measure induced from a
state by the Born rule? The answer is yes by Gleason's theorem when
the dimension~$\ge3$~\cite{gleason1957,peres1995quantum,Redhead1987-REDINA}.
Furthermore, a simple corollary of Gleason's theorem can show the
Born rule is the unique function satisfying conditions 1. to 3.
\begin{cor}
The Born rule is the unique function satisfying conditions 1. to 3.
\end{cor}
\begin{proof}
Assume there is another function $\tilde{\mu}'$ such that $\tilde{\mu}'_{\rho}$
is a quantum probability measure for all mixed state~$\rho$. We
are going to prove $\tilde{\mu}'=\tilde{\mu}$.

>>>>

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\newcommand{\yutsung}[1]{\fbox{\begin{minipage}{0.9\textwidth}\color{purple}{Yu-Tsung says: #1}\end{minipage}}}
\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.9\textwidth}\color{green}{Amr says: #1}\end{minipage}}}
\newcommand{\ffzd}[1]{{\mathbb{F}^{d\;*}_{#1}}}
\def\C{{\mathbb{C}}}
\newcommand{\ff}[1]{\mathbb{F}_{#1}}
\newcommand{\melement}[2]{ \langle #1 | #2 | #1 \rangle}
\newcommand{\expect}[2]{ \langle #1 | #2 | #1 \rangle}

% The following copy from `iopart.cls'
\newcommand{\Tr}{\mathop{\mathrm{Tr}}\nolimits}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\rme}{\mathrm{e}}

@@

\begin{example}\label{One-qubit-quantum-probability}[One-qubit quantum
probability space] Consider a one-qubit Hilbert space with states
  $\alpha \ket{0} + \beta \ket{1}$ such that
  $|\alpha|^2 + |\beta|^2 = 1$. The set of events associated with this
  Hilbert space consists of all projection operators. Each event is
  interpreted as a possible post-measurement state of a quantum system
  in current state $\ket{\phi}$. For example, the event $\proj{0}$
  indicates that the post-measurement state will be~$\ket{0}$; the
  event $\proj{1}$ indicates that the post-measurement state will
  be~$\ket{1}$; the event $\proj{\ps}$ where
  $\ket{\ps} = \frac{1}{\sqrt{2}}(\ket{0}+\ket{1})$ indicates that the
  post-measurement state will be $\ket{\ps}$; the event
  $\mathbb{1} = \proj{0}+\proj{1}$ indicates that the post-measurement
  state will be a linear combination of $\ket{0}$ and $\ket{1}$; and
  the empty event $\mathbb{0}$ states that the post-measurement state
  will be the empty state. As in the classical case, a probability
  measure is a function that maps events to $[0,1]$: here is a partial
  specification of a possible probability measure:
\[\begin{array}{rcl}
\mu\left(\mathbb{0}\right) = 0, \quad
\mu\left(\mathbb{1}\right) =  1, \quad
\mu\left(\proj{0}\right) = 1, \quad
\mu\left(\proj{1}\right) = 0, \quad
\mu\left(\proj{\ps}\right) = 1/2, \quad \ldots
\end{array}\]
Note that, similarly to the classical case, the probability of
$\mathbb{1}$ is 1 and the probability of collections of orthogonal
events (e.g., $\proj{0}+\proj{1}$) is the sum of the individual
probabilities. In contrast, a collection of non-orthogonal events
(e.g., $\proj{0}$ and $\proj{\ps}$) is not itself an event.
In the classical example, we argued that each probability measure is
uniquely determined by two actual coins. A similar (but much more
subtle) argument is valid also in the quantum case. By postulates of
quantum mechanics and Gleason's theorem, it turns out that for large
enough quantum systems, each probability measure is uniquely
determined by an actual quantum state.
\qed\end{example}

@@

For a given set of events~$\events$, there are many possible probability
measures~$\mu:\events\rightarrow[0,1]$. The Born rule~\cite{Born1984,Mermin2007},
a postulate of quantum mechanics, states that each pure quantum state
$\ket{\phi}\in\Hilb\backslash\left\{ 0\right\} $ induces a probability
measure~$\mu_{\phi}$ as follows: 
\[
\mu_{\phi}(E)=\frac{\ip{\phi}{E\phi}}{\ip{\phi}{\phi}}
\]
This Born rule can be extended to mixed quantum states~$\rho$ by
\begin{eqnarray}
\mu_{\rho}\left(E\right) & = & \Tr\left(\rho E\right)=\sum_{j=1}^{N}q_{j}\mu_{\psi_{j}}\left(E\right)\textrm{ ,}\label{BornRule.mixed}
\end{eqnarray}
where $\rho=\sum_{j=1}^{N}q_{j}\frac{\proj{\psi_{j}}}{\ip{\psi_{j}}{\psi_{j}}}$
is a density matrix, and $\sum_{j=1}^{N}q_{j}=1$, i.e., $\Tr\left(\rho\right)=1$.
It is easy to verify that given a mixed state, the Born rule also
induces a probability measure. Conversely, Gleason's theorem states
that given a probability measure~$\mu$, there exist a mixed quantum
states~$\rho$ that induces such a measure using the Born rule. The
theorem is only valid in Hilbert spaces with dimension $d\geq3$.
It is instructive to study counterexamples in $d=2$, i.e., the case
of a one-qubit system. As in example~\ref{One-qubit-quantum-probability},
the set of events associated with it contains $\mathbb{0}$, $\mathbb{1}$,
and one-dimensional projection operators which can be expressed as
$\proj{\phi}$, where $\ket{\phi}=\left(\begin{array}{c}
\cos\theta\\
\rme^{\rmi\gamma}\sin\theta
\end{array}\right)$. Then, we can define a probability measure as follow.
\begin{eqnarray*}
\mu\left(\mathbb{0}\right) & = & 0\textrm{;}\\
\mu\left(\mathbb{1}\right) & = & 1\textrm{;}\\
\mu\left(\proj{\phi}\right) & = & \begin{cases}
0 & \textrm{, if }0\le\theta<\frac{\pi}{2}\textrm{;}\\
1 & \textrm{, if }\frac{\pi}{2}\le\theta<\pi\textrm{.}
\end{cases}
\end{eqnarray*}
$\mu$ is a probability measure, but not induced by the Born rule
because $\mu$ not continuous, but the probabilities induces by the
Born rule are always continuous.

\yutsung{Rieder and Svozil~\cite{RiederSvozil2007} didn't provide
enough details to construct an example. It seems that the full details
in written in Svozil's book~\cite{Svozil1998}. Hence, I have recommended
the library to purchase this book and sent an interlibrary loan request
as well.}

@@

Although it seems that we need an infinite long table to specify the
quantum probability measure~$\mu$, our $\mu$ is actually
given by a simple formula~$\melement{0}{E}$. In general, Born discovered
each quantum state $\ket{\psi}\in\Hilb\backslash\left\{ 0\right\} $
induces a probability measure $\tilde{\mu}_{\psi}:\qevents\rightarrow[0,1]$
on the space of events defined for any event $E\in\qevents$ as follows~\cite{Born1984,Mermin2007}:
\begin{equation}
\tilde{\mu}_{\psi}(E)=\frac{\melement{\psi}{E}}{\ip{\psi}{\psi}}\label{eq:Born}
\end{equation}
The Born rule satisfies the following properties:
\begin{enumerate}
\item It can be extend to mixed states. Given a mixed state represented
by a density matrix $\rho=\sum_{j=1}^{N}q_{j}\frac{\proj{\psi_{j}}}{\ip{\psi_{j}}{\psi_{j}}}$,
where $\sum_{j=1}^{N}q_{j}=1$, i.e., $\Tr\left(\rho\right)=1$, then
the Born rule can be extended to 
\begin{eqnarray}
\mu_{\rho}\left(E\right) & = & \Tr\left(\rho E\right)=\sum_{j=1}^{N}q_{j}\mu_{\psi_{j}}\left(E\right)\textrm{ ,}\label{BornRule.mixed-1}
\end{eqnarray}
where $\rho=\sum_{j=1}^{N}q_{j}\frac{\proj{\psi_{j}}}{\ip{\psi_{j}}{\psi_{j}}}$
is a density matrix, and $\sum_{j=1}^{N}q_{j}=1$, i.e., $\Tr\left(\rho\right)=1$.
\\
Notice that $\left(\left\{ 1,\ldots,N\right\} ,2^{\left\{ 1,\ldots,N\right\} },\pmeas\left(J\right)=\sum_{j\in J}q_{j}\right)$
is a classical probability space. Therefore, when we discretize the
Hilbert space later, we may need to discretize this probability space
as well. 
\item $\tilde{\mu}_{\rho}$ is a probability measure for all mixed state~$\rho$. 
\item $\ip{\psi}{\phi}=0\Leftrightarrow\tilde{\mu}_{\psi}\left(\proj{\phi}\right)=0$. 
\item $\tilde{\mu}_{\psi}\left(E\right)=\tilde{\mu}_{\mathbf{U}\ket{\psi}}\left(\mathbf{U}E\mathbf{U}^{\dagger}\right)\textrm{ ,}$where
$\mathbf{U}$ is any unitary map, i.e., $\mathbf{U}^{\dagger}\mathbf{U}=\mathbb{1}$. 
\end{enumerate}
Naturally, we may ask: is every probability measure induced from a
state by the Born rule? The answer is yes by Gleason's theorem when
the dimension~$\ge3$~\cite{gleason1957,peres1995quantum,Redhead1987-REDINA}.
Furthermore, a simple corollary of Gleason's theorem can show the
Born rule is the unique function satisfying conditions 1. to 3.
\begin{cor}
The Born rule is the unique function satisfying conditions 1. to 3.
\end{cor}
\begin{proof}
Assume there is another function $\tilde{\mu}'$ such that $\tilde{\mu}'_{\rho}$
is a quantum probability measure for all mixed state~$\rho$. We
are going to prove $\tilde{\mu}'=\tilde{\mu}$.
